Contextual recruitment of selective attention can be updated via changes in task-relevance

### 2019-09-30 manuscript_pretty.Rmd
almost there

In Experiment 1, we found that

@@

Experiment 2, therefore, serves as a control experiment to test these two assumptions. Experiment 2 is identical to Experiment 1 with two important changes: First, we removed the counting task, giving no explicit instructions about how to use the context cues; We only informed participants that images would be present throughout the task. This allowed us to determine if there was any systematic effect of biased items on the unbiased items absent the counting task. Second, we altered the proportion congruency of the biased item sets from 100%/0% to 75%/25%. This allowed us to measure the congruency effect for the biased item sets to determine whether participants learned about the associations between proportion congruency and the context cues without the counting task [e.g., @crump_flexibility_2009]. If this were the case, we would expect larger congruency effects for the 75% proportion congruent items as compared to the 25% proportion congruent items.

@@

In the current study, we have bene examining the importance of task-relevance for the contextual recruitment of selective attention. In Experiment 1, we found that task-relevance allowed participants to exploit context cues in the presence of a competing, contradictory cue and subsequently ignore a previously learned association in favor of a different context cue. In Experiment 2, we conceptually replicated previous work showing that task-relevance was necessary for participants to learn about the biased proportion congruency associations [e.g., @crump_flexibility_2009] and ruled out possible confounding effects of the biased items.

Experiment 3

@@

Importantly, participants were given further instructions to exploit the proportion congruency associated with context cues.

@@

The results of the RT analysis revealed the critical, two-way interaction between task-relevant context and unbiased-item congruency to be non-significant, `r RT_ANOVA_E3$Task_Relevant_Context_Congruency`. Similarly, the corresponding error analysis also resulted in a non-significant two-way interaction between the task-relevant context and unbiased-item congruency, `r ACC_ANOVA_E3$Task_Relevant_Context_Congruency`.

# Estimating reproducability by simulation-based sensitivity- and power-analyses

@@

(ref:figure5) Results from Experiment 3 showing congruency effects (incongruent - congruent) for frequency unbiased items in reaction times (left) and error rates (right) as a function of the task-relevant context (100% proportion congruent versus 0% proportion congruent). Task-relevance was manipulated through explicit instructions.

>>>>

In Experiment 1, we found that manipulating the task-relevance of a shared contextual feature was associated in predictable changes in context-specific proportion congruency effects for frequency unbiased context items; Namely, larger congruency effects when the context was associated with low conflict than when the context was associated with high conflict.  This suggests that participants learned the association between context and conflict-level in the first block (despite multiple, competing cues), and subsequently ignored that association in the second block to learn a new association. Task-relevance then, may not just be important for establishing new associations [e.g.,@crump_flexibility_2009] but also important for determining which learned associations are expressed in the present moment.

@@

Experiment 2, therefore, serves as a control experiment to test these two assumptions directly. Experiment 2 is identical to Experiment 1 with two important changes: First, we removed the counting task, giving no explicit instructions about how to use the context cues; We only informed participants that images would be present throughout the task. This allowed us to determine if there was any systematic effect of biased items on the unbiased items absent the counting task. Second, we altered the proportion congruency of the biased item sets from 100%/0% to 75%/25%. This allowed us to measure the congruency effect for the biased item sets to determine whether participants learned about the associations between proportion congruency and the context cues without the counting task [e.g., @crump_flexibility_2009]. If this were the case, we would expect larger congruency effects for the 75% proportion congruent items as compared to the 25% proportion congruent items.

@@

In the current study, we have been examining the importance of task-relevance for the contextual recruitment of selective attention. In Experiment 1, we found that task-relevance allowed participants to exploit context cues in the presence of a competing, contradictory cue and subsequently ignore a previously learned association in favor of a different context cue. In Experiment 2, we conceptually replicated previous work showing that task-relevance was necessary for participants to learn about the biased proportion congruency associations [e.g., @crump_flexibility_2009] and ruled out possible confounding effects of the biased items.

Experiment 3 tests the boundaries of the task-relevance effect found in Experiment 1. The evidence thus far suggests that context-cues, rather than being irrelevant to the task (as their namesake would suggest), need to be integrated into the task-representation to guide attentional control. There is some evidence that instructions alone can be sufficient to produce list-wide proportion congruency effects, though not item-specific proportion congruency effects (e.g., Entel et al., 2014; though, see Desender, 2018) and not context-specific proportion congruency effects (e.g., ). Explicit instructions however, are another way of making context-cues relevant to the on-going task and could serve the same purpose as the counting task---although, instructions are likely a much weaker manipulation than the counting task. In Experiment 3, we test whether participants are able to explicitly use the context-cues to guide attentional control through an instruction manipulation.

@@

Importantly, participants were given further instructions to exploit the proportion congruency associated with context cues. Halfway through the experiment, participants received new instructions about which context features were relevant and associated with high/low conflict. For example:

> *"On every trial you'll see five arrows (pointed left or right) paired with an image.*

> *The images will help you perform this task.*

> *The arrows will either be congruent (the center arrow is the same as the flanking arrows) or incongruent (the center arrow is different from the flanking arrows). Incongruent trials are typically harder because you need to ignore the flanking arrows, while the congruent trials are easier, because you do not need to ignore the flanking arrows.*

> *In the first block of the experiment, trials that include a chair (regardless of whether it is blue or green) will be mostly incongruent. That is, most of the trials that include a chair will be hard.*

> *This will be helpful to perform the task, so try to proactively use this information when you perform the task."*

@@

The results of the RT analysis revealed the critical, two-way interaction between task-relevant context and unbiased-item congruency to be non-significant, `r RT_ANOVA_E3$Task_Relevant_Context`. Similarly, the corresponding error analysis also resulted in a non-significant two-way interaction between the task-relevant context and unbiased-item congruency, `r ACC_ANOVA_E3$Task_Relevant_Context`.

# Estimating replicability by simulation-based sensitivity- and power-analyses

@@

(ref:figure5) Results from the simulation-based sensitivity and power-analyses. Figure A shows the sensitivity of our experimental design to detect a range of effect sizes for the context-specific proportion congruency (CSPC) effect with 50 (e.g., Experiments 2 and 3) and 150 (e.g., Experiment 1) participants. Figure B shows the sensitivity to detect the CSPC by context-type interaction with 50 participants per group (e.g., Experiment 1). Figures C and D show the estimated power curves for the CSPC and CSPC by context-type effects across sample sizes. Each data point represents the proportion of p-values less than .05 from 10,000 simulations.

In Experiment 1, we found no evidence for differences in the task-relevance effect across context-type. Similarly, Experiments 2 and 3 failed to find CSPC effects. To effectively draw inferences from these results, we need to assess the extent to which our experiments could have detected effects, if present. Although replicability has become an important and salient issue in Psychology at large (e.g., ), this issue has also arisen in the context-specific proportion congruency literature (e.g., Brosowsky Crump, Hutcheon speiler, TiCS).

To address this issue in the current study---and to do so transparently---we conducted simulation-based power and sensitivity analyses (Maxwell, Kelley, and Raush, 2008; Brosowsky and Crump) to determine the range effect sizes our experimental design was sensitive to detect and the required sample sizes to detect them. We used a Monte Carlo simulation approach with a statistical model of the distributions that underlie the effect of interest (Maxwell, 2008). Rather than estimating power based on observed effect sizes from prior experiments (typically, over-estimates), we instead estimated the properties of the base reaction time distributions from our current data and sampled from these distributions using a Monte Carlo simulation approach to estimate the replication success rate across a range of effect sizes and sample sizes.

We created the conflictPower R-package to conduct the Monte Carlo simulations (Crump and Brosowsky; Tics). The conflictPower package samples simulated reaction time data for subjects at the trial-level, given base reaction time distributions for the conflict and no-conflict conditions. To generate the base reaction time distributions, we fitted ex-Gaussian functions to the reaction time data from individual participants and averaged across all three experiments using the retimes R-package (XXXX). We then used the conflictPower package to simulate the results across a range of effect and sample sizes, calculating the proportion of experiments, across $10 000$ simulations, resulting in a p-value less than $.05$ (see Figure 5; for a more detailed explanation of the sampling procedure see Crump and Brosowsky, 2019).

First, the results of these analyses show that our experimental design was reasonably powered to detect CSPC effects. In particular, with 150 participants, we estimated 89% power to detect effects as small 10 ms; With only 50 participants, we were still reasonably powered to detect CSPC effects as small as 15 ms at an estimated 78%. Understandably, our design was not as sensitive to detecting changes the in the CSPC effect across conditions (in Experiment 1, for example). We estimated only 75% power to detect the difference between a 20 ms and 0 ms CSPC effect with 50 participants per group. That is, the experimental design could detect the presence/absence of the CSPC effect moderately well, but likely does not reliably detect small changes in the CSPC effect across conditions. To reliably detect a 15 ms change in the CSPC effect at roughly 80% power, for example, we estimate would require 100 participants per group.

### 2019-09-30 manuscript_pretty.tex
almost there

In Experiment 1, we found that

We made two assumptions in the design of Experiment 1: First, we assumed that task-relevance was required for participants to learn about the associations between context cues and the biased proportion congruency manipulations. Our task was inspired by Crump and Milliken (2009) who failed find context-specific effects using a task-irrelevant shape cue until they used a counting task to make the shapes relevant. However, location is the most-often used context cue in these kinds of paradigms (e.g., ). In these cases, no task-relevance manipulation is necessary to produce CSPC effects using location cues. Though, it is arguable whether location is truly task-irrelevant when localizing the target is necessary for responding. Regardless, it is unclear in our paradigm whether participants required the counting task to learn about the biased proportion congruency manipulations.

@@

Experiment 2, therefore, serves as a control experiment to test these two assumptions. Experiment 2 is identical to Experiment 1 with two important changes: First, we removed the counting task and gave no explicit instructions about how to use the context cues; We only informed participants that images would be present throughout the task. This allowed us to determine if there was any systematic effect of biased items on the unbiased items absent the counting task. Second, we altered the proportion congruency of the biased item sets from 100\%/0\% to 75\%/25\%. This allowed us to measure the congruency effect for the biased item sets to determine whether participants learned about the associations between proportion congruency and the context cues without the counting task (e.g., Crump \& Milliken, 2009). If this were the case, we would expect larger congruency effects for the 75\% proportion congruent items as compared to the 25\% proportion congruent items.

@@

In the current study, we have been homing in on the importance of task-relevance for the contextual recruitment of selective attention. In Experiment 1, we found that task-relevance allowed participants to exploit context cues in the presence of a competing, contradictory cue and subsequently ignore the previously learned association in favor of a different context cue. In Experiment 2, we conceptually replicated previous work showing that task-relevance was necessary for participants to learn about the biased proportion congruency associations (e.g., Crump \& Milliken, 2009) and ruled out possible confounding effects of the biased items.

@@

Importantly, participants were given further instructions to exploit the proportion congruency associated with context cues.

@@

The results of the RT analysis revealed the critical, two-way interaction between task-relevant context and unbiased-item congruency to be non-significant, . Similarly, the corresponding error analysis also resulted in a non-significant two-way interaction between the task-relevant context and unbiased-item congruency, .

\hypertarget{estimating-reproducability-by-simulation-based-sensitivity--and-power-analyses}{%
\section{Estimating reproducability by simulation-based sensitivity- and power-analyses}\label{estimating-reproducability-by-simulation-based-sensitivity--and-power-analyses}}

\begin{figure}
\centering
\includegraphics{manuscript_pretty_files/figure-latex/figure5-1.pdf}
\caption{\label{fig:figure5}Results from Experiment 3 showing congruency effects (incongruent - congruent) for frequency unbiased items in reaction times (left) and error rates (right) as a function of the task-relevant context (100\% proportion congruent versus 0\% proportion congruent). Task-relevance was manipulated through explicit instructions.}
\end{figure}

>>>>

In Experiment 1, we found that manipulating the task-relevance of a shared contextual feature was associated in predictable changes in context-specific proportion congruency effects for frequency unbiased context items; Namely, larger congruency effects when the context was associated with low conflict than when the context was associated with high conflict. This suggests that participants learned the association between context and conflict-level in the first block (despite multiple, competing cues), and subsequently ignored that association in the second block to learn a new association. Task-relevance then, may not just be important for establishing new associations (e.g., Crump \& Milliken, 2009) but also important for determining which learned associations are expressed in the present moment.

We made two assumptions in the design of Experiment 1: First, we assumed that task-relevance was required for participants to learn about the associations between context cues and the biased proportion congruency manipulations. Our task was inspired by Crump and Milliken (2009) who failed find context-specific effects using a task-irrelevant shape cue until they used a counting task to make the shapes relevant. However, location is the most-often used context cue in these kinds of paradigms (e.g., ). In these cases, no task-relevance manipulation is necessary to produce CSPC effects, though, it is arguable whether location is truly task-irrelevant when localizing the target is necessary for responding. Regardless, it is unclear in our paradigm whether participants required the counting task to learn about the biased proportion congruency manipulations.

@@

Experiment 2, therefore, serves as a control experiment to test these two assumptions directly. Experiment 2 is identical to Experiment 1 with two important changes: First, we removed the counting task, giving no explicit instructions about how to use the context cues; We only informed participants that images would be present throughout the task. This allowed us to determine if there was any systematic effect of biased items on the unbiased items absent the counting task. Second, we altered the proportion congruency of the biased item sets from 100\%/0\% to 75\%/25\%. This allowed us to measure the congruency effect for the biased item sets to determine whether participants learned about the associations between proportion congruency and the context cues without the counting task (e.g., Crump \& Milliken, 2009). If this were the case, we would expect larger congruency effects for the 75\% proportion congruent items as compared to the 25\% proportion congruent items.

@@

In the current study, we have been examining the importance of task-relevance for the contextual recruitment of selective attention. In Experiment 1, we found that task-relevance allowed participants to exploit context cues in the presence of a competing, contradictory cue and subsequently ignore a previously learned association in favor of a different context cue. In Experiment 2, we conceptually replicated previous work showing that task-relevance was necessary for participants to learn about the biased proportion congruency associations (e.g., Crump \& Milliken, 2009) and ruled out possible confounding effects of the biased items.

Experiment 3 tests the boundaries of the task-relevance effect found in Experiment 1. The evidence thus far suggests that context-cues, rather than being irrelevant to the task (as their namesake would suggest), need to be integrated into the task-representation to guide attentional control. There is some evidence that instructions alone can be sufficient to produce list-wide proportion congruency effects, though not item-specific proportion congruency effects (e.g., Entel et al., 2014; though, see Desender, 2018) and not context-specific proportion congruency effects (e.g., ). Explicit instructions however, are another way of making context-cues relevant to the on-going task and could serve the same purpose as the counting task---although, instructions are likely a much weaker manipulation than the counting task. In Experiment 3, we test whether participants are able to explicitly use the context-cues to guide attentional control through an instruction manipulation.

@@

Importantly, participants were given further instructions to exploit the proportion congruency associated with context cues. Halfway through the experiment, participants received new instructions about which context features were relevant and associated with high/low conflict. For example:

\begin{quote}
\emph{"On every trial you'll see five arrows (pointed left or right) paired with an image.}
\end{quote}

\begin{quote}
\emph{The images will help you perform this task.}
\end{quote}

\begin{quote}
\emph{The arrows will either be congruent (the center arrow is the same as the flanking arrows) or incongruent (the center arrow is different from the flanking arrows). Incongruent trials are typically harder because you need to ignore the flanking arrows, while the congruent trials are easier, because you do not need to ignore the flanking arrows.}
\end{quote}

\begin{quote}
\emph{In the first block of the experiment, trials that include a chair (regardless of whether it is blue or green) will be mostly incongruent. That is, most of the trials that include a chair will be hard.}
\end{quote}

\begin{quote}
\emph{This will be helpful to perform the task, so try to proactively use this information when you perform the task."}
\end{quote}

@@

The results of the RT analysis revealed the critical, two-way interaction between task-relevant context and unbiased-item congruency to be non-significant, \(F(1, 45) = 0.02\), \(\mathit{MSE} = 4,030.54\), \(p = .876\), \(\hat{\eta}^2_p = .001\), 90\% CI \([0\), \(0.04]\). Similarly, the corresponding error analysis also resulted in a non-significant two-way interaction between the task-relevant context and unbiased-item congruency, \(F(1, 45) = 0.66\), \(\mathit{MSE} = 28.52\), \(p = .420\), \(\hat{\eta}^2_p = .014\), 90\% CI \([0\), \(0.11]\).

\hypertarget{estimating-replicability-by-simulation-based-sensitivity--and-power-analyses}{%
\section{Estimating replicability by simulation-based sensitivity- and power-analyses}\label{estimating-replicability-by-simulation-based-sensitivity--and-power-analyses}}

\begin{figure}
\centering
\includegraphics{manuscript_pretty_files/figure-latex/figure5-1.pdf}
\caption{\label{fig:figure5}Results from the simulation-based sensitivity and power-analyses. Figure A shows the sensitivity of our experimental design to detect a range of effect sizes for the context-specific proportion congruency (CSPC) effect with 50 (e.g., Experiments 2 and 3) and 150 (e.g., Experiment 1) participants. Figure B shows the sensitivity to detect the CSPC by context-type interaction with 50 participants per group (e.g., Experiment 1). Figures C and D show the corresponding power curves for the CSPC and CSPC by context-type effects across sample sizes.}
\end{figure}

In Experiment 1, we found no evidence for differences in the task-relevance effect across context-type. Similarly, Experiments 2 and 3 failed to find CSPC effects. To effectively draw inferences from these results, we need to assess the extent to which our experiments could have detected effects, if present. Although replicability has become an important and salient issue in Psychology at large (e.g., ), this issue has also arisen in the context-specific proportion congruency literature (e.g., Brosowsky Crump, Hutcheon speiler, TiCS).

To address this issue in the current study---and to do so transparently---we conducted simulation-based power and sensitivity analyses (Maxwell, Kelley, and Raush, 2008; Brosowsky and Crump) to determine the range effect sizes our experimental design was sensitive to detect and the required sample sizes to detect them. We used a Monte Carlo simulation approach with a statistical model of the distributions that underlie the effect of interest (Maxwell, 2008). Rather than estimating power based on observed effect sizes from prior experiments (typically, over-estimates), we instead estimated the properties of the base reaction time distributions from our current data and sampled from these distributions using a Monte Carlo simulation approach to estimate the replication success rate across a range of effect sizes and sample sizes.

We created the conflictPower R-package to conduct the Monte Carlo simulations (Crump and Brosowsky; Tics). The conflictPower package samples simulated reaction time data for subjects at the trial-level, given base reaction time distributions for the conflict and no-conflict conditions. To generate the base reaction time distributions, we fitted ex-Gaussian functions to the reaction time data from individual participants and averaged across all three experiments using the retimes R-package (XXXX). Using the extimated ex-Gaussian parameters, we used the conflictPower package to simulate the results across a range of effect and sample sizes, calculating the proportion of experiments (across \(10 000\) simulations) resulting in a p-value less than \(.05\) (see Figure 5; for a more detailed explanation of the sampling procedure see Crump and Brosowsky, 2019).

First, the results of these analyses show that our experimental design was reasonably powered to detect CSPC effects. In particular, with 150 participants, we estimated 89\% power to detect effects as small 10 ms; With only 50 participants, we were still reasonably powered to detect CSPC effects as small as 15 ms at an estimated 78\%. Understandably, our design was not as sensitive to detecting changes the in the CSPC effect across conditions (in Experiment 1, for example). For instance, we estimated only 75\% power to detect the difference between a 20 ms and 0 ms CSPC effect with 50 participants per group. That is, the experimental design could detect the presence/absence of the CSPC effect moderately well, but likely does not reliably detect small changes in the CSPC effect across conditions. To reliably detect a 15 ms change in the CSPC effect, for example, we estimate would require 100 participants per group.
